---
title: "Project 1 Team 8"
# date: "today"
#date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  #pdf_document:
    #toc: yes
    #toc_depth: '3'
---

```{r init, include=F}
library(ezids)
```


```{r setup, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
knitr::opts_chunk$set(warning = F, results = "markup", message = F)
# knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. 
```

```{r}
p<-read.csv("D:\\Dataset1.csv")
str(p)
summary(p)
p$ethnicity<-as.factor(p$ethnicity)
p$gender<-as.factor(p$gender)
p$gender<-as.factor(p$icu_type)
summary(p)

p <- p[, !names(p) %in% c('bilirubin_apache')]
p <- p[, !names(p) %in% c('creatinine_apache')]
p <- p[, !names(p) %in% c('glucose_apache')]
str(p)
```

```{r}
p[p == ''] <- NA
s=na.omit(p)
#summary(p)
s[1:10,]
#summary(s)
nrow(s)
```

```{r}
library(ggplot2)
ggplot(data.frame(x=s$heart_rate_apache), aes(x))+geom_histogram(binwidth = 10, color="black",fill="blue")+labs(xlab="heart rate apache", ylab="frequency",title="visualization of heart rate with outliers")
ggplot(s, aes(x = heart_rate_apache), y = "") +
  geom_boxplot() +
  labs(title = "Box Plot of heart rate",
       x = "heart rate apache", y = "frequency distribution") +
  theme_minimal()
qqnorm(s$heart_rate_apache, main = "heart rate Q-Q Plot", ylab="heart rate", datax=FALSE)
qqline(s$heart_rate_apache, col = "red")
```









```{r}
c <- s$heart_rate_apache

q1 <- quantile(c, 0.25)
q3 <- quantile(c, 0.75)

iqr_value <- IQR(c)

lower_bound <- q1 - 1.5 * iqr_value
upper_bound <- q3 + 1.5 * iqr_value


outliers <- c < lower_bound | c > upper_bound


print("Outliers:")
print(c[outliers])

print("Lower Bound:")
print(lower_bound)
print("Upper Bound:")
print(upper_bound)

cleaned_data <- s[!(s$heart_rate_apache %in% outliers$heart_rate_apache),]
nrow(cleaned_data)

```

```{r}
#library(ggplot2)
#ggplot(data.frame(x=cleaned_data$heart_rate_apache), aes(x))+geom_histogram(binwidth = 2, #color="black",fill="blue")+labs(title="visualization of heart rate without outliers")

#ggplot(data.frame(x=cleaned_data$bmi), aes(x))+geom_histogram(binwidth = 2, color="black",fill="blue")+labs(title="visualization of #bmi with outliers")
```

```{r}
c3 <- cleaned_data$bmi


q1bmi <- quantile(c3, 0.25)
q3bmi <- quantile(c3, 0.75)


iqr_valuebmi <- IQR(c3)


lower_boundbmi <- q1bmi - 1.5 * iqr_valuebmi
upper_boundbmi <- q3bmi + 1.5 * iqr_valuebmi


outliersbmi <- cleaned_data[c3 < lower_boundbmi | c3 > upper_boundbmi, ]


print("Outliers:")
print(outliersbmi)


print("Lower Bound:")
print(lower_boundbmi)
print("Upper Bound:")
print(upper_boundbmi)




cleaned_data3 <-cleaned_data[!(cleaned_data$bmi %in% outliersbmi$bmi),]
#nrow(s)
nrow(cleaned_data3)

```

```{r}
c4 <- cleaned_data3$age


q1age <- quantile(c4, 0.25)
q3age <- quantile(c4, 0.75)


iqr_valueage <- IQR(c4)


lower_boundage <- q1age - 1.5 * iqr_valueage
upper_boundage <- q3age + 1.5 * iqr_valueage


outliersage <- cleaned_data3[c4 < lower_boundage | c4 > upper_boundage, ]


print("Outliers:")
print(outliersage)


print("Lower Bound:")
print(lower_boundage)
print("Upper Bound:")
print(upper_boundage)




cleaned_data4 <-cleaned_data3[!(cleaned_data3$age %in% outliersage$age),]
#nrow(s)
nrow(cleaned_data4)

```

```{r}
ggplot(cleaned_data4, aes(x = ethnicity, y = bmi, fill = ethnicity)) +
 geom_col() +
  labs(title = "Bar Plot considering ethnicity and bmi",
       x = "ethnicity", y = "bmi")


ggplot(cleaned_data4, aes(x = icu_type, fill = factor(hospital_death))) +
  geom_bar(position = "dodge") +
  labs(title = "Bar Plot of icu type and hospital death",
       x = "icu_type", y = "Count") +
  theme_minimal()


# Box Plot
ggplot(cleaned_data4, aes(x = factor(cirrhosis), y = heart_rate_apache)) +
  geom_boxplot() +
  labs(title = "Box Plot of cirrhosis and heart rate",
       x = "cirrhosis", y = "heart_rate_apache") +
  theme_minimal()

# Violin Plot
ggplot(cleaned_data4, aes(x = factor(cirrhosis), y = heart_rate_apache, fill = factor(cirrhosis))) +
  geom_violin() +
  labs(title = "Violin Plot of cirrhosis and heart rate",
       x = "cirrhosis", y = "heart_rate_apache") +
  theme_minimal()


# Box Plot
ggplot(cleaned_data4, aes(x = factor(hepatic_failure), y = heart_rate_apache)) +
  geom_boxplot() +
  labs(title = "Box Plot of hepatic failure and heart rate",
       x = "hepatic_failure", y = "heart_rate_apache") +
  theme_minimal()

# Violin Plot
ggplot(cleaned_data4, aes(x = factor(hepatic_failure), y = heart_rate_apache, fill = factor(hepatic_failure))) +
  geom_violin() +
  labs(title = "Violin Plot of hepatic failure and heart rate",
       x = "hepatic_failure", y = "heart_rate_apache") +
  theme_minimal()


# Box Plot
ggplot(cleaned_data4, aes(x = factor(diabetes_mellitus), y = heart_rate_apache)) +
  geom_boxplot() +
  labs(title = "Box Plot of diabetes and heart rate",
       x = "diabetes_mellitus", y = "heart_rate_apache") +
  theme_minimal()

# Violin Plot
ggplot(cleaned_data4, aes(x = factor(diabetes_mellitus), y = heart_rate_apache, fill = factor(diabetes_mellitus))) +
  geom_violin() +
  labs(title = "Violin Plot of diabetes and heart rate",
       x = "diabetes_mellitus", y = "heart_rate_apache") +
  theme_minimal()


# Box Plot
ggplot(cleaned_data4, aes(x = factor(leukemia), y = heart_rate_apache)) +
  geom_boxplot() +
  labs(title = "Box Plot of leukemia and heart rate",
       x = "leukemia", y = "heart_rate_apache") +
  theme_minimal()

# Violin Plot
ggplot(cleaned_data4, aes(x = factor(leukemia), y = heart_rate_apache, fill = factor(leukemia))) +
  geom_violin() +
  labs(title = "Violin Plot of leukemia and heart rate",
       x = "leukemia", y = "heart_rate_apache") +
  theme_minimal()

# Box Plot
ggplot(cleaned_data4, aes(x = factor(lymphoma), y = heart_rate_apache)) +
  geom_boxplot() +
  labs(title = "Box Plot of lymphoma vs heart rate",
       x = "lymphoma", y = "heart_rate_apache") +
  theme_minimal()

# Violin Plot
ggplot(cleaned_data4, aes(x = factor(lymphoma), y = heart_rate_apache, fill = factor(lymphoma))) +
  geom_violin() +
  labs(title = "Violin Plot of lymphoma vs heart rate",
       x = "lymphoma", y = "heart_rate_apache") +
  theme_minimal()
```

```{r}




# Box Plot
ggplot(cleaned_data4, aes(x = factor(cirrhosis), y = bmi)) +
  geom_boxplot() +
  labs(title = "Box Plot cirrhosis and bmi" ,
       x = "cirrhosis", y = "bmi") +
  theme_minimal()

# Violin Plot
ggplot(cleaned_data4, aes(x = factor(cirrhosis), y = bmi, fill = factor(cirrhosis))) +
  geom_violin() +
  labs(title = "Violin Plot cirrhosis and bmi",
       x = "cirrhosis", y = "bmi") +
  theme_minimal()


# Box Plot
ggplot(cleaned_data4, aes(x = factor(hepatic_failure), y = bmi)) +
  geom_boxplot() +
  labs(title = "Box Plot hepatic failure and bmi",
       x = "hepatic_failure", y = "bmi") +
  theme_minimal()

# Violin Plot
ggplot(cleaned_data4, aes(x = factor(hepatic_failure), y = bmi, fill = factor(hepatic_failure))) +
  geom_violin() +
  labs(title = "Violin Plot hepatic failure and bmi",
       x = "hepatic_failure", y = "bmi") +
  theme_minimal()


# Box Plot
ggplot(cleaned_data4, aes(x = factor(diabetes_mellitus), y = bmi)) +
  geom_boxplot() +
  labs(title = "Box Plot diabetes_mellitus and bmi",
       x = "diabetes_mellitus", y = "bmi") +
  theme_minimal()

# Violin Plot
ggplot(cleaned_data4, aes(x = factor(diabetes_mellitus), y = bmi, fill = factor(diabetes_mellitus))) +
  geom_violin() +
  labs(title = "Violin Plot diabetes_mellitus and bmi",
       x = "diabetes_mellitus", y = "bmi") +
  theme_minimal()


# Box Plot
ggplot(cleaned_data4, aes(x = factor(leukemia), y = bmi)) +
  geom_boxplot() +
  labs(title = "Box Plot leukemia and bmi",
       x = "leukemia", y = "bmi") +
  theme_minimal()

# Violin Plot
ggplot(cleaned_data4, aes(x = factor(leukemia), y = bmi, fill = factor(leukemia))) +
  geom_violin() +
  labs(title = "Violin Plot leukemia and bmi",
       x = "leukemia", y = "bmi") +
  theme_minimal()

# Box Plot
ggplot(cleaned_data4, aes(x = factor(lymphoma), y = bmi)) +
  geom_boxplot() +
  labs(title = "Box Plot lymphoma and bmi",
       x = "lymphoma", y = "bmi") +
  theme_minimal()

# Violin Plot
ggplot(cleaned_data4, aes(x = factor(lymphoma), y = bmi, fill = factor(lymphoma))) +
  geom_violin() +
  labs(title = "Violin Plot lymphoma and bmi",
       x = "lymphoma", y = "bmi") +
  theme_minimal()



ggplot(cleaned_data4, aes(x=heart_rate_apache,y=bmi, color=age))+geom_point()+labs(title="heart rate vs bmi color coded based on age")
```

```{r}
contingency_table <- table(cleaned_data4$cirrhosis, cleaned_data4$hospital_death)
print(contingency_table)

chi_square_result <- chisq.test(table(cleaned_data4$hospital_death, cleaned_data4$cirrhosis))
print(chi_square_result)

contingency_table1 <- table(cleaned_data4$leukemia, cleaned_data4$hospital_death)
print(contingency_table1)

chi_square_result1 <- chisq.test(table(cleaned_data4$hospital_death, cleaned_data4$leukemia))
print(chi_square_result1)

contingency_table2 <- table(cleaned_data4$lymphoma, cleaned_data4$hospital_death)
print(contingency_table2)

chi_square_result2 <- chisq.test(table(cleaned_data4$hospital_death, cleaned_data4$lymphoma))
print(chi_square_result2)

contingency_table3 <- table(cleaned_data4$diabetes_mellitus, cleaned_data4$hospital_death)
print(contingency_table3)

chi_square_result3 <- chisq.test(table(cleaned_data4$hospital_death, cleaned_data4$diabetes_mellitus))
print(chi_square_result3)

contingency_table4 <- table(cleaned_data4$hepatic_failure, cleaned_data4$hospital_death)
print(contingency_table4)

chi_square_result4 <- chisq.test(table(cleaned_data4$hospital_death, cleaned_data4$hepatic_failure))
print(chi_square_result4)
```

```{r}
atest1<-aov(heart_rate_apache ~ icu_type,data=cleaned_data4)
print(summary(atest1))
library(TukeyC)
hoctest<-TukeyHSD(atest1)
hoctest
```



```{r}
library(caret)
encoding <- dummyVars("~ .", data = cleaned_data4)
encoded_data <- predict(encoded_data_model, newdata = cleaned_data4)
#print(encoded_data)
summary(encoded_data)
str(encoded_data)
m=as.data.frame(encoded_data)
#print(m)
str(m)
```

```{r}
install.packages("e1071")
library(e1071)

skew_before <- skewness(m$bmi)
print(skew_before)
#Log transformation
transformed_data <- log(m$bmi + 1)  
print(skewness(transformed_data))
m$bmi<-transformed_data
```

```{r}
group_0 <- m$heart_rate_apache[m$cirrhosis == 0]
group_1 <- m$heart_rate_apache[m$cirrhosis == 1]

t_test_result <- t.test(group_0, group_1)

print(t_test_result)
```


```{r}
#install.packages("corrplot")  
library(corrplot)
cor_matrix <- cor(m[, c("heart_rate_apache", "bmi", "age")])
print(cor_matrix)
corrplot(cor_matrix, method = "color",  type = "upper", 
         #col = col(200),     
         tl.col = "black",  
         tl.srt = 90,         
         order = "hclust",    
         addCoef.col = "black", 
         number.cex = 0.5,  
         title = "Correlation Matrix", mar=c(0,0,0,0))
```








# Question 1: 
**Among different age groups (children, adults, and senior citizens), which demographic exhibits the highest mortality rate?**
```{r}
data$AgeGroup <- ifelse(data$age < 18, "Children", ifelse(data$age < 65, "Adults", "Senior Citizens"))

# Calculate the mortality rate for each age group
mortality_table <- table(data$AgeGroup, data$hospital_death)

# Calculate the mortality rates
mortality_rate <- (mortality_table[, "1"] / rowSums(mortality_table)) * 100

# Create a dataframe for the results
mortality_results <- data.frame(
  AgeGroup = names(mortality_rate),
  MortalityRate = mortality_rate
)

# Find the demographic with the highest mortality rate
highest_mortality_group <- mortality_results[which.max(mortality_results$MortalityRate), ]

barplot(mortality_results$MortalityRate, names.arg = mortality_results$AgeGroup, col = "skyblue",
        xlab = "Age Group", ylab = "Mortality Rate (%)", main = "Mortality Rate by Age Group")

# Add text to indicate the group with the highest mortality rate
text(x = which.max(mortality_results$MortalityRate), y = highest_mortality_group$MortalityRate + 1,
     label = paste("Highest:", round(highest_mortality_group$MortalityRate, 2), "%"), col = "red")

cat("The demographic with the highest mortality rate is", highest_mortality_group$AgeGroup,
    "with a mortality rate of", highest_mortality_group$MortalityRate, "%.")

```
# Question 2:
**Can the presence of medical conditions like diabetes, hepatic failure, or other medical conditions impact an individual's likelihood of mortality?**
```{r}
data$diabetes_mellitus <- as.factor(data$diabetes_mellitus)
data$hepatic_failure <- as.factor(data$hepatic_failure)

model <- glm(hospital_death ~ diabetes_mellitus + hepatic_failure, data = data, family = "binomial")

summary(model)

# Chi-squared test for diabetes_mellitus
chisq_test_diabetes <- chisq.test(table(data$diabetes_mellitus, data$hospital_death))
print(chisq_test_diabetes)

# Chi-squared test for hepatic_failure
chisq_test_hepatic <- chisq.test(table(data$hepatic_failure, data$hospital_death))
print(chisq_test_hepatic)
```
# Question 3:
**What is the gender distribution of deaths in the hospital, with regard to male and female ratios? ** 
```{r}
gender_table <- table(data$gender, data$hospital_death)

# Extract the counts of deceased males and females
deceased_males <- gender_table["M", "1"]
deceased_females <- gender_table["F", "1"]

# Calculate the male-to-female ratio
male_to_female_ratio <- deceased_males / deceased_females



cat("The male-to-female ratio among deceased individuals in the hospital is:", male_to_female_ratio)
```
# Question 4:
**Among ethnicities, which one exhibits the highest and lowest mortality rate?**
```{r}

library(dplyr)


mortality_rates <- data %>%
  group_by(ethnicity) %>%
  summarise(
    total_patients = n(),
    total_deaths = sum(hospital_death),
    mortality_rate = (total_deaths / total_patients) * 100  
  )


highest_mortality <- mortality_rates[which.max(mortality_rates$mortality_rate), ]
lowest_mortality <- mortality_rates[which.min(mortality_rates$mortality_rate), ]


cat("Ethnicity with Highest Mortality Rate: ", highest_mortality$ethnicity, "\n")
cat("Highest Mortality Rate: ", highest_mortality$mortality_rate, "%\n")
cat("Ethnicity with Lowest Mortality Rate: ", lowest_mortality$ethnicity, "\n")
cat("Lowest Mortality Rate: ", lowest_mortality$mortality_rate, "%\n")

```

